---
title: "Gibi Ratings"
author: "Ozge Gundogan"
date: "2026-09-13"
format: html
---

# Introduction

This project will examine rating data for the first six episodes of the television series "Gibi". Ratings will be collected from five users using a 1-5 scale and organized into a dataset that can be stored and analyzed using SQL and R.

# Planned Approach

First, rating data will be collected from five users for the first six episodes of "Gibi". Users will rate only the episodes they have watched. If a user has not watched an episode, the rating will be recorded as "NULL" in the SQL database. The collected data will then be stored in a PostgreSQL database using separate tables for users, episodes, and ratings. Finally, the data will be loaded into R as a dataframe, where the number of ratings and average rating will be calculated for each episode. Missing ratings will be excluded from the average-rating calculations.

# Anticipated Data Challenges

One anticipated challenge is ensuring that the data are transferred correctly from the SQL database into R as a dataframe. The SQL query will need to retrieve the correct user, episode, and rating information, while missing values will need to be handled appropriately in R.

### Load Required Packages

The analysis will use "DBI" and "RPostgres" to connect R to the PostgreSQL database. The "dplyr" package will be used to inspect and summarize the data after it is loaded into R.

```{r required_packages}
library(DBI)
library(RPostgres)
library(dplyr)
library(rstudioapi)
library(ggplot2)
```

### Connect to the PostgreSQL Database

This section establishes a connection between R and the PostgreSQL database. The connection allows the analysis to access the stored users, episodes, and ratings data directly from R.

```{r database_connection}
#| eval: false
con <- dbConnect(
  RPostgres::Postgres(),
  dbname = "DATA607",
  host = "localhost",
  port = 5432,
  user = "postgres",
  password = rstudioapi::askForPassword("Enter PostgreSQL password")
)
```

### Create the Database Tables

The database uses separate tables for users, episodes, and ratings. This structure separates the entities and allows the ratings table to connect users with the episodes they rated.

``` sql
CREATE TABLE users (
    user_id SERIAL PRIMARY KEY,
    name VARCHAR(100) NOT NULL
);
```

``` sql
CREATE TABLE episodes (
    episode_id SERIAL PRIMARY KEY,
    episode_number INTEGER NOT NULL,
    title VARCHAR(100) NOT NULL
);
```

``` sql
CREATE TABLE ratings (
    rating_id SERIAL PRIMARY KEY,
    user_id INTEGER NOT NULL REFERENCES users(user_id),
    episode_id INTEGER NOT NULL REFERENCES episodes(episode_id),
    rating INTEGER CHECK (rating >= 1 AND rating <= 5)
);
```

``` sql
INSERT INTO users (name)
VALUES
    ('Serpil'),
    ('Koray'),
    ('Okan'),
    ('Deniz'),
    ('Pelin');
```

``` sql
SELECT * FROM users;
```

``` sql
INSERT INTO episodes (episode_number, title)
VALUES
    (1, 'Kokaric'),
    (2, 'Vatka'),
    (3, 'Nu Model'),
    (4, 'Erasmusla GY'),
    (5, 'Yanlis Mentor'),
    (6, 'Karanlik Guc');
```

``` sql
SELECT * FROM episodes;
```

``` sql
INSERT INTO ratings (user_id, episode_id, rating)
VALUES
    (1, 1, 5),
    (1, 2, 4),
    (1, 3, 4),
    (1, 4, 5),
    (1, 5, NULL),
    (1, 6, 5),

    (2, 1, 5),
    (2, 2, 5),
    (2, 3, 4),
    (2, 4, 5),
    (2, 5, 3),
    (2, 6, 5),

    (3, 1, 5),
    (3, 2, 4),
    (3, 3, NULL),
    (3, 4, 5),
    (3, 5, 3),
    (3, 6, 5),

    (4, 1, 5),
    (4, 2, 4),
    (4, 3, NULL),
    (4, 4, 5),
    (4, 5, NULL),
    (4, 6, 4),

    (5, 1, 4),
    (5, 2, 5),
    (5, 3, 4),
    (5, 4, 5),
    (5, 5, 3),
    (5, 6, 5);
```

``` sql
SELECT * FROM ratings;
```

``` sql
SELECT *
FROM ratings
WHERE rating IS NULL;
```

``` sql
SELECT
    u.name, 
    e.episode_number,
    e.title,
    r.rating
FROM ratings r
JOIN users u ON r.user_id = u.user_id
JOIN episodes e ON r.episode_id = e.episode_id
ORDER BY e.episode_number, u.name;
```

### Load Data from PostgreSQL

This section retrieves the rating data from the PostgreSQL database and loads it into R as a dataframe. The SQL query combines the users, episodes, and ratings tables using joins so that each rating is associated with the corresponding user and episode.

```{r load_from_postgresql}
#| eval: false
gibi_ratings <- dbGetQuery(
  con, "
  select
    u.name,
    e.episode_number,
    e.title,
    r.rating
  from ratings r
  join users u on r.user_id = u.user_id
  join episodes e on r.episode_id = e.episode_id
  order by e.episode_number, u.name;
  "
)
gibi_ratings
```

### Load Data from GitHub

The rating data are also available as a CSV file in the GitHub repository. This allows the analysis to be reproduced without requiring access to the local PostgreSQL database.

```{r load_from_github}
gibi_ratings <- read.csv("https://raw.githubusercontent.com/OzgeG01/DATA607-Assignment/refs/heads/main/gibi_ratings.csv")
gibi_ratings
```

### Inspect the Data Structure

This section checks the structure of the dataframe to confirm the number of observations and variables and to verify that the data were imported with the expected data types.

```{r}
str(gibi_ratings)
```

### Check for Missing Data

This section checks the dataframe for missing rating values. Missing ratings represent episodes that a participant did not rate, so they are excluded when calculating rating counts and average ratings.

```{r missing_count}
sum(is.na(gibi_ratings$rating))
```

```{r missing_data}
missing_data <- gibi_ratings %>% 
  filter(is.na(rating))

missing_data
```

### Ratings per User

This section counts the number of valid ratings provided by each user. Missing ratings are excluded from the count because they represent episodes that a user did not rate.

```{r ratings_per_user}
ratings_per_user <- gibi_ratings %>%
  group_by(name) %>%
  summarise(
    number_of_ratings = sum(!is.na(rating)),
    .groups = "drop"
  )
ratings_per_user
```

### Ratings per Episode

This section counts the number of valid ratings received by each episode. Missing ratings are excluded from the count.

```{r ratings_per_episode}
ratings_per_episode <- gibi_ratings %>%
  group_by(episode_number, title) %>%
  summarise(
    number_of_ratings = sum(!is.na(rating)),
    .groups = "drop"
  )
ratings_per_episode
```

### Average Ratings

This section calculates the average rating for each episode. Missing ratings are excluded from the calculation.

```{r average_ratings}
average_ratings <- gibi_ratings %>%
  group_by(episode_number, title) %>%
  summarise(
    average_rating = mean(rating, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(average_rating))
average_ratings
```

The average ratings can be compared using the following bar chart.

```{r average_rating_plot}
ggplot(average_ratings, aes(x = reorder(title, average_rating),
                            y = average_rating)) + 
  geom_col() +
  coord_flip() + 
  labs(
    title = "Average Ratings by Gibi Episode",
    x = "Episode",
    y = "Average Rating"
  ) + 
  theme_minimal()
```

# Conclusions

The analysis successfully loaded the "Gibi" episode ratings from PostgreSQL into R and summarized the available ratings by participant and episode. Missing ratings were identified and excluded from the rating counts and average rating calculations.

The results show that "Erasmusla GY" received the highest average rating of 5.0, while "Yanlis Mentor" received the lowest average rating of 3.0. "Kokaric" and "Karanlik Guc" both received an average rating of 4.8, followed by "Vatka" with 4.4 and "Nu Model" with 4.0.

As a recommendation, the analysis could be extended by collecting ratings from more participants to make the results more representative. Future analysis could also compare individual participant preferences and examine whether the episode rankings change as additional ratings are collected.
